---
title: "PCA Imputation"
output:
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# installation of packages in case the user has not installed yet
requirements <- c("nnet", "MASS", "e1071", "class", "leaps", "glmnet", "car", "caTools",
              "pROC", "summarytools", "dplyr", "ggplot2", "tidyverse", "lubridate",
              "mapview", "sf", "geojsonio", "leaflet", "broom", "gridExtra",
              "effects")

for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}

library(mgcv)
```


```{r}
source("00_set_up.R")

na_counts <- colSums(is.na(data))
na_counts <- na_counts[na_counts > 0]

for (nm in names(na_counts)) {
  cat(nm, ":", na_counts[nm], "\n")
}

```

# Preprocessing and Feature Engineering 

First, the dataset was split into training (2020) and test (2024).
The date field was handled as a proper Date, and a month feature was engineered as a 12-level factor (“01”–“12”), which is useful both for seasonality in models and for grouped imputation.

There are some missing values in no2_mean, o3_mean, so2_mean, co_mean and their AQI variants. Instead of being arbitrarily eliminated, missing values were then examined and filled using PCA.


A missingness GLM (missing indicator ~ city + month + temp + humidity + windspeed) suggested patterned missingness for NO₂/O₃/SO₂ (associated to city/month/temp/humidity/windspeed), while CO missingness looked random.

The project employed a straightforward, comprehensible imputation: PCA imputation calculated on the training set, subsequently used to both train and test because patterned missingness can skew outcomes if we delete rows.

As only the mean  concentrations of the pollutants are used in building the final model, the missing values are filled in only those columns.


## Dealing with Missing Values

```{r }

## Fit GLM for Missing Values

vars <- c("no2_mean","o3_mean", "so2_mean", "co_mean")
preds <- c("city","month","temp","humidity","windspeed")

for (v in vars) {

  miss_name <- paste0("miss_", v)
  data[[miss_name]] <- as.integer(is.na(data[[v]]))

  dat <- data[, c(miss_name, preds)]
  dat <- dat[complete.cases(dat), ]

  fit <- glm(as.formula(paste(miss_name, "~", paste(preds, collapse=" + "))),
             data = dat, family = binomial)

  cat("\n============================\n")
  cat("Missingness model for:", v, " (n =", nrow(dat), ")\n")
  cat("============================\n")

  print(summary(fit))   # <-- this includes stars automatically
}



```
The missing values of NO2 shows a pattern with city Monroe and humidity, it also have some relationship with the predictor windspeed.

On the other hand, o3_mean has a pattern with cities Erie, Queens, and Bronx. It shares the relationship with month10 and month12.

The missing values of so2 also demonstrates a pattern with cities Monroe and Queens, and a little bit with temperature.

The previous analysis tells that co_mean missing value is random and doesnot have any pattern, thus, doesnot share any relationship with other predictors. 


Therefore, NO₂ / O₃ / SO₂ missingness is patterned (city/month/weather effects), so dropping all rows with any of those missing can introduce bias (and also removes more data). So I imputed using train-based city×month medians and optionally included missingness indicators.

### Applying PCA for Imputation
```{r}

library(missMDA)

# include variables that help explain pollutants

vars_pca <- c("co_mean","no2_mean","so2_mean","o3_mean",
              "pm2_5_mean","temp","humidity","windspeed")


vars_pca <- vars_pca[vars_pca %in% names(train_all)]

Xtr <- data.matrix(train_all[, vars_pca])
Xte <- data.matrix(test_all[,  vars_pca])

# number of principal components
ncp <- estim_ncpPCA(Xtr, ncp.max = 6)$ncp

# impute TRAIN
imp_tr <- imputePCA(Xtr, ncp = ncp, scale = TRUE)
train_all[, vars_pca] <- imp_tr$completeObs

# impute TEST (separately)
imp_te <- imputePCA(Xte, ncp = ncp, scale = TRUE)
test_all[, vars_pca] <- imp_te$completeObs

# check remaining NA (should be 0)
colSums(is.na(train_all[, vars_pca]))
colSums(is.na(test_all[,  vars_pca]))


```



```{r}

saveRDS(train_all, "train_all_2020_impute.rds")
saveRDS(test_all,  "test_all_2024_impute.rds")

# later
train_full <- readRDS("train_all_2020_impute.rds")
test_full  <- readRDS("test_all_2024_impute.rds")

```

```{r}
# first 6 rows
head(train_full)
```

```{r}
colSums(is.na(train_full))
colSums(is.na(test_full))
```

```{r}
na_counts <- colSums(is.na(train_full))
na_counts <- na_counts[na_counts > 0]

for (nm in names(na_counts)) {
  cat(nm, ":", na_counts[nm], "\n")
}
```
After the imputation of the missing  values, I have build the first baseline Linear regression model, Please go to the Part 3 - **lm.Rmd**  to continue the analysis.